# Philosophies of Data

## What are Data

Data are observations about the world. In order to make sense of and give
context to these observations we organize data into structures.

## Data Frames and Tibbles

R is an object oriented programming language, and there are several varieties
of R objects you can use to represent data. In this book we will concentrate
mostly on how to use data frames as the central object for thinking about data.
Data frames are rectangular data objects, meaning that they are composed of
rows and columns, much like the layout of a piece of graph paper or a
spreadsheet. In the next few chapters we'll discuss how you can import data
into R from many different sources, but you can also create data frames with
code alone.

There's no better tool for creating your own data frames from scratch than the
`tibble` package. Install the package with `install.packages("tibble")`, and
then we can start creating our own data frames.

Let's start up the package and then create a simple data frame, where every
column is specified with a simple vector:

```{r}
library(tibble)

people <- data_frame(Name = c("Shannon", "Jeff", "Sean", "Roger"),
           Department = c("Genetics", "Biostatistics", "Cognitive Science", "Biostatistics"),
           School = c("JHU", "JHU", "UCSD", "Monash"))
people
```

Let's take a look at what is printed when we enter our table of `people` in the
R console. The first line, `# A tibble: 4 x 3`, shows that the object being
printed *is* a tibble. A tibble is a data frame with some nice extra properties
which we'll discuss later. The second line, `Name   Department   School`, shows
the column names for the tibble. The third line, `<chr>   <chr>   <chr>`, shows
the type of data in each column. After the third line, each line shows one row
in the dataset, starting with a row number.

We can add rows and columns using the `add_row()` and `add_column()` functions
respectively. The `add_row()` function takes an argument for each column in the
tibble while the `add_column()` function takes an argument for each row in the
tibble.

```{r}
people <- add_row(people, Name = "Brian", Department = "Biostatistics", School = "JHU")
people

people <- add_column(people, Puppet = c(FALSE, TRUE, FALSE, TRUE, TRUE))
people
```

## Tidy Data

## Matrix Data

## Hierarchical Data